hl|tc|Groups, Rings, and Fields
\( \text{Definition 2.1.} \) A Group is a set \(G\) equipped with a binary operation \( (\cdot) : G \times G \rightarrow G \) that associates an element \( a \cdot b \in G \ \forall \ a, b \in G \) such that,
li| \( \ a \cdot (b \cdot c) = (a \cdot b) \cdot c \ \ \forall a,b,c \in G \)         (G1 : Associativity)
li| \( \exists e \mid a \cdot e = a = e \cdot a \ \ \forall a \in G  \)                  (G2 : Existance of Identity)
li| \( \forall a \in G, \ \exists a^{-1} \mid a \cdot a^{-1} = e = a^{-1} \cdot a \)     (G3 : Existance of Inverse)
A group is called commutative (or abelian) if \( a \cdot b = b \cdot a \ \ \forall a,b \in G \).
A set \(M\) with an operation \( (\cdot) : M \times M \rightarrow M \) which satisfies (G1,G2) is called a Monoid.
\( \text{Example 2.1.}\)
li| \( (Z, +) \) is a commutative group, but \( (Z, \times) \) is a monoid.
li| \( (Q, +) \) & \( (Q - \{0\}, \times) \) are commutative groups.
li| The set of bijections (permutations) \( f : S \rightarrow S \) & composition operation \( (\circ) \) is called the Symmetric Group of degree \(|S|\) (& order \(n!\)).
li| \( (Z/mZ, +) \) is a commutative group, & the set of integers coprime to \(m\) form a commutative group under mutliplication which is often represented as \( (Z/mZ)^\times \) 
li| The set of invertible \(n \times n\) metrices forms a group under matrix multiplication, which is known as general linear group, \( GL(n, R) \), \( GL(n, C) \), etc. When restricted to matrices with determinant \(1\), the structure is called special linear group, \( SL(n, R) \), \( SL(n, C) \), etc. Linear groups of matrices \(Q\) with real coefficients such that \( Q Q^T = Q^T Q = I \) are called general & special orthogonal groups & are represented by \( O(n), SO(n) \) respectively.
It is customary to denote abelian group operation by \(+\) & inverse as \( -a \) for any element \( a \in G\).
\( \text{Proposition 2.1.} \) For a Monoid \( (M, \cdot) \), left-identity must be the same as right-identity.
tc| \( (e' \cdot a = a \ \forall \ a \in M) \land (a \cdot e'' = a \ \forall \ a \in M) \implies e' = e' \cdot e'' = e'' \)
\( \text{Proposition 2.2.} \) For a Monoid \( (M, \cdot) \), left-inverse must be the same as right-inverse, if either exists for a given element.
tc| \( (a' \cdot a = e) \land (a \cdot a'' = e) \implies a' = a' \cdot e = a' \cdot (a \cdot a'') = (a' \cdot a) \cdot a'' = e \cdot a'' = a'' \)
\( \text{Proposition 2.3.} \) For a Monoid \( (M, \cdot) \), if we have two invertible elements \(\{a,b\}\), then \( (a \cdot b) \) is also invertible.
tc| \( (a \cdot b) \cdot (b^{-1} \cdot a^{-1}) = a \cdot (b \cdot b^{-1}) \cdot a^{-1} = e \implies (a \cdot b)^{-1} = b^{-1} \cdot a^{-1} \)